---
title: "Handling violated assumptions - linear regression & alternatives"
author: "Mike B."
date: "11/10/2025"
format: revealjs
---

```{r}
#| include: false
#| label: setup

#standard set
library(tidyverse)
library(emmeans)
library(marginaleffects)

#for bayesian modeling
#install.packages(c("brms", "bayestestR", "bayesplot", "tidybayes"))
library(brms)
library(bayestestR)
library(bayesplot)
library(tidybayes)

palette <- viridis::rocket(10)

```

## When assumptions aren't met

Sometimes our data doesn't meet the assumptions of the analytical tools we would like to use - in my experience, more often than not. We often need to choose a different approach to make sense of our data, and we will look at a few options related to skewed data when we have linear regression in mind for our analysis. 

## A note on planning analysis

I usually recommend researchers don't plan the specific analytical tool they plan to apply (i.e., linear regression, t-test, etc.). Instead, I advise they plan on the effects they want to describe and choose the appropriate analysis when the data is known - especially unbounded continuous data.

## Data

This data is from a fairly recent experiment, where we ended up with with skewed data.

```{r}
#| label: data
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
frac <- read_csv("data/Fracture_GcGf_v1.csv") |> 
  mutate(
    treatment =
      case_when(
        treatment == "control" ~ "Control",
        treatment == "70C" ~ "70°C",
        treatment == "140C" ~ "140°C"
      ),
    treatment = fct_relevel(factor(treatment), "Control", "70°C", "140°C"),
    orientation = if_else(orientation == "R", "TR", "RT")
  )

frac |> slice_head(n = 10)
```

## Visualise

```{r}
#| label: vis_raw
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Linear regression

```{r}
#| label: fit_lm
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Adjusting with a log transformation

```{r}
#| label: fit_log_transform
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Predictions

```{r}
#| label: predict_each_val_gf
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

```{r}
#| label: predict_each_val_gf_log
#| echo: true
#| code-fold: true
#| code-summary: "click for code"


```

## Comparisons

## Standard linear regression

```{r}
#| label: compare_lm
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Log-transformed

```{r}
#| label: compare_log
#| echo: true
#| code-fold: true
#| code-summary: "click for code"


```

## Visualising

```{r}
#| label: vis_mods
#| echo: true
#| code-fold: true
#| code-summary: "click for code"

```

## Gamma regression

```{r}
#| label: gamma_regression
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Another alternative - Bayesian regression


## Packages

```{r}
#| label: bayes_packages
#| echo: true
#| code-fold: true
#| code-summary: "click for code"

#loaded previously, but repeated here for informative purposes
# library(brms)
# library(bayestestR)
# library(bayesplot)
# library(tidybayes)

```


## Fitting a bayesian model requires estimating prior knowledge

```{r}
#| label: bayes_priors
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Bayesian regression fitting

```{r}
#| label: bayes_fit
#| echo: true
#| code-fold: true
#| code-summary: "click for code"

```

## Output

```{r}
#| label: bayes_out
#| echo: true
#| code-fold: true
#| code-summary: "click for code"

```

## A closer look: EPI:140C

```{r}
#| label: bayes_est_vs_lm_est
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

