---
title: "Homework 3 - DSP1 - 2025/2026"
author: "Dushime Mudahera Richard 76250043"
toc: true
format: html
---

Homework 3

Step 1: Rename this file as follows: "DSP1-HW3-Name-StudentID" with your name and student ID replacing the appropriate parts.
Step 2: Update the author field in the YAML as indicated.
Step 3: Ensure the document has a Table of Contents when rendered (hint: you set this in the YAML)
Step 3: Complete to the Tasks below. Be sure to add meaningful label tags to each R code chunk and structure your analysis sensibly.
Step 4: Render to html or pdf (you may need to change target formats in the YAML)
Step 5: Zip (or otherwise compress) the html other relevant support folders and include this qmd file with your work filled in
Step 6: Submit the compressed folder which should be named "DSP1-HW3-Name-StudentID" through the e-classroom by Tuesday 6 January 2026.

Hint: Make sure that when I open the html file I can view all the embedded images, see all your code and easily identify your responses to questions.

Note: Please see the rules for using AI on the e-classroom.

## Ordinal regression and beta regression

This homework will cover two regression types: logistic and ordinal regression. 

## Ordinal regression

A recent survey asked respondents to rate wood wall panels they viewed on a several semantic differential scales. These scales use two adjectives with opposite meanings and ask respondents to describe some stimuli by selecting a point on the scale between the two adjectives.

In this case, the scale was between artificial and natural with the following ordered categories:

Artificial, Somewhat artificial, Neutral, Somewhat Natural, Natural

The data set "artificial_natural.csv" includes a numeric version of this scale (-2 to 2) and a character version.

Fit a an ordinal regression model that uses the character version of the scale ("labs" in the data provided) as the dependent variable and Panel (8 level categorical), Age (numeric, continuous), and NR.mu (numeric, continuous) as the predictor variables. Use a _logit_ link function. "labs" will need to be converted to an ordered factor with the factors in the correct order as above. The other variables are self explanatory enough, but NR.mu is the mean score of the items on the nature relatedness scale [NR-6 version](https://www.frontiersin.org/journals/psychology/articles/10.3389/fpsyg.2013.00813/full).

  1. Are there differences in ratings between panel types? Provide a text explanation and visualisation that illustrates the
  differences on the probability scale.
  2. What is the effect of age? Is it meaningful? Provide a text explanation and a visualisation to explain the outcome.
  3. Does nature relatedness have an effect? Provide a text explanation and a visualisation to explain the outcome.
  4. Create a visualisation of the boundaries between artificial-natural scale categories overlaid on a logistic distribution for the
  2waycladding panel (baseline unless you change something) and the singlebrown panel (two separate visualisations are ok) In each,
  highlight the probability of selecting "somewhat natural" by using a fill between the boundaries for that category and label of the
  probability of selecting that category.
  5. Refit the model using a _probit_ link. State which model you prefer between the logit and probit model and why. (Refer to the Bürkner and Vuorre (2019) reading on the e-classroom for support.)

## setup

```{r}
#| label: setup
library(tidyverse)
library(ordinal)
library(viridis)
library(emmeans)
palette <- viridis(10, option = "magma")
`%notin%` <- Negate(`%in%`)
```

### Loading and Preparing Data

```{r}
#| label: load-data

d.an <- read_csv("data/artificial_natural.csv")

an_levels <- c("Artificial", "Somewhat artificial", "Neutral", "Somewhat natural", "Natural")
d.an <- d.an %>%
  mutate(labs = factor(labs, levels = an_levels, ordered = TRUE))

```
```{r}
summary(d.an)
```

## Fitting the Model

Uses the `clm` function from the `ordinal` package with a logit link.

```{r}
#| label: fit-logit
model.logit <- clm(labs ~ Panel + Age + NR.mu, data = d.an, link = "logit")
summary(model.logit)
```

### Differences in Ratings Between Panel Types

To assess differences, we examine the coefficients for each Panel level relative to the baseline (2waycladding). Significant p-values indicate differences. Positive coefficients suggest a shift towards more "Natural" ratings.

```{r}
#| label: panel-probs
newdata_panel <- expand_grid(Panel = unique(d.an$Panel), 
                             Age = mean(d.an$Age, na.rm = TRUE), 
                             NR.mu = mean(d.an$NR.mu, na.rm = TRUE))
preds_panel <- predict(model.logit, newdata = newdata_panel, type = "prob")$fit
newdata_panel <- cbind(newdata_panel, preds_panel) %>%
  pivot_longer(cols = all_of(an_levels), names_to = "Category", values_to = "Probability") %>%
  mutate(Category = factor(Category, levels = an_levels))

ggplot(newdata_panel, aes(x = Panel, y = Probability, fill = Category)) +
  geom_col(position = "stack") +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  labs(title = "Predicted Probabilities by Panel Type", y = "Probability", x = "Panel") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Panels such as "logwall" and "clt" have significant positive coefficients, indicating they are perceived as more natural compared to the baseline "2waycladding".

For "singlebrown" also shows a positive effect . Conversely, "osb" and "oakverticals" have significant negative coefficients, suggesting more artificial perceptions.

For "brownverticals" and "cleavedpattern"  are not significantly different from the baseline. 

The visualization shows probability distributions across categories for each panel, holding Age and NR.mu at their means.


### Effect of Age

The coefficient for Age is 0.0163 with p-value 7.19e-11, indicating a significant positive effect. Older participants tend to rate panels as more natural.

Is it meaningful? The effect size is small (0.016 per year), but statistically significant, suggesting a gradual shift towards more natural perceptions with increasing age.

```{r}
#| label: age-effect
newdata_age <- expand_grid(Age = seq(min(d.an$Age, na.rm = TRUE), max(d.an$Age, na.rm = TRUE), length.out = 50),
                           Panel = "2waycladding",  # Baseline
                           NR.mu = mean(d.an$NR.mu, na.rm = TRUE))
preds_age <- predict(model.logit, newdata = newdata_age, type = "prob")$fit
newdata_age <- cbind(newdata_age, preds_age) %>%
  pivot_longer(cols = all_of(an_levels), names_to = "Category", values_to = "Probability") %>%
  mutate(Category = factor(Category, levels = an_levels))

# plotting 
ggplot(newdata_age, aes(x = Age, y = Probability, color = Category, group = Category)) +
  geom_line() +
  scale_color_viridis_d(option = "magma") +
  theme_minimal() +
  labs(title = "Predicted Probabilities Across Age", y = "Probability")
```

### Effect of Nature Relatedness (NR.mu)

The coefficient for NR.mu is 0.307 with p-value 5.53e-09, showing a significant positive effect. Higher nature relatedness leads to more "Natural" ratings.

```{r}
#| label: nr-effect
newdata_nr <- expand_grid(NR.mu = seq(min(d.an$NR.mu, na.rm = TRUE), max(d.an$NR.mu, na.rm = TRUE), length.out = 50),
                          Panel = "2waycladding",
                          Age = mean(d.an$Age, na.rm = TRUE))
preds_nr <- predict(model.logit, newdata = newdata_nr, type = "prob")$fit
newdata_nr <- cbind(newdata_nr, preds_nr) %>%
  pivot_longer(cols = all_of(an_levels), names_to = "Category", values_to = "Probability") %>%
  mutate(Category = factor(Category, levels = an_levels))

ggplot(newdata_nr, aes(x = NR.mu, y = Probability, color = Category, group = Category)) +
  geom_line() +
  scale_color_viridis_d(option = "magma") +
  theme_minimal() +
  labs(title = "Predicted Probabilities Across NR.mu", y = "Probability")
```

### Visualization of Boundaries

For the logit link, we plot the logistic density with thresholds. Highlight "Somewhat natural" (between tau[3] and tau[4]).

For 2waycladding (baseline, no shift).

```{r}
#| label: boundaries-baseline
tau <- model.logit$alpha
latent_df <- tibble(x = seq(-6, 6, length.out = 500)) %>%
  mutate(y = dlogis(x))
prob_sn <- plogis(tau[4]) - plogis(tau[3])

ggplot(latent_df, aes(x = x, y = y)) +
  geom_line() +
  geom_vline(xintercept = tau, linetype = "dashed", color = "red") +
  geom_area(data = filter(latent_df, x > tau[3] & x < tau[4]), aes(y = y), fill = "skyblue", alpha = 0.5) +
  annotate("text", x = (tau[3] + tau[4])/2, y = 0.05, label = paste("Somewhat natural\nProb:", round(prob_sn, 3))) +
  theme_minimal() +
  labs(title = "Latent Scale for 2waycladding", x = "Latent Variable", y = "Density")
```

For singlebrown: Shift by its coefficient.

```{r}
#| label: boundaries-singlebrown
coef_sb <- coef(model.logit)["Panelsinglebrown"]
latent_df_sb <- tibble(x = seq(-6, 6, length.out = 500)) %>%
  mutate(y = dlogis(x), x_shifted = x - coef_sb)  # Shift left if positive coef (towards more natural)
prob_sn_sb <- plogis(tau[4] - coef_sb) - plogis(tau[3] - coef_sb)

ggplot(latent_df_sb, aes(x = x_shifted, y = y)) +
  geom_line() +
  geom_vline(xintercept = tau - coef_sb, linetype = "dashed", color = "red") +
  geom_area(data = filter(latent_df_sb, x_shifted > tau[3] - coef_sb & x_shifted < tau[4] - coef_sb), aes(y = y), fill = "skyblue", alpha = 0.5) +
  annotate("text", x = (tau[3] + tau[4])/2 - coef_sb, y = 0.05, label = paste("Somewhat natural\nProb:", round(prob_sn_sb, 3))) +
  theme_minimal() +
  labs(title = "Latent Scale for singlebrown", x = "Latent Variable", y = "Density")
```

### Probit Model and Comparison

```{r}
#| label: fit-probit
model.probit <- clm(labs ~ Panel + Age + NR.mu, data = d.an, link = "probit")
summary(model.probit)
```

```{r}
#| label: model-comparison
AIC(model.logit)
AIC(model.probit)
```

I prefer the Logit model. Not only is the AIC lower (6248.22 vs 6253.99), indicating a better statistical fit, but the logit link is also more interpretable in psychology. Following Bürkner and Vuorre (2019), the logit model allows us to interpret the coefficients as Odds Ratios. For example, the nature relatedness (NR.mu) estimate of 0.307 implies that for every 1-unit increase in nature relatedness, the odds of a respondent moving to a 'more natural' category increase by approximately 36% ($e^{0.307} \approx 1.36$


## Beta regression

Use the attached data about democracy worldwide. It is a subset from the (Varieties of Democracy)[https://www.v-dem.net] project (V-Dem). You will model the proportion of female (prop_fem) members of parliament (MP) with different predictors.

The data is in the "vdem_2020_subset.csv" file.

1. Visualise the proportion of women in parliament based on quota in two different ways.
2. Fit a beta regression with prop_fem as the response variable with quota and polyarchy (degree of democracy) as predictor variables.
3. Visualise the proportion of women MPs based on quota
4. Visualise the proportion of women MPs based on polyarchy
5. Write a brief explanation of the outcomes, refer to previous visualisations or create new ones to help with the explanation.

### setup

```{r}
#| label: setup-beta
library(tidyverse)
library(betareg)
library(emmeans)
library(marginaleffects)
library(viridis)

palette <- viridis(10, option = "turbo")
inv.logit <- function(x) { exp(x) / (1 + exp(x)) }
```

### loading data 

```{r}
#| label: load-data-beta
d.vdem <- read_csv("data/vdem_2020_subset.csv")
```
```{r}
summary(d.vdem)
```

### Proportion of Women in Parliament by Quota

Way 1: Boxplot

```{r}
#| label: boxplot-quota
ggplot(d.vdem, aes(x = factor(quota, labels = c("No", "Yes")), y = prop_fem)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(title = "Boxplot: Prop Fem by Quota", x = "Quota", y = "Proportion Female MPs")
```

Way 2: Density Plot

```{r}
#| label: density-quota
ggplot(d.vdem, aes(x = prop_fem, fill = factor(quota, labels = c("No", "Yes")))) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Density: Prop Fem by Quota", x = "Proportion Female MPs", fill = "Quota")
```

### Fit Beta Regression Model

```{r}
#| label: fit-beta
model.beta <- betareg(prop_fem ~ quota + polyarchy, data = d.vdem)
summary(model.beta)
```

### Proportion by Quota

Predicted proportions for each quota level, holding polyarchy at mean.

```{r}
#| label: preds-quota
preds_quota <- predictions(model.beta, by = "quota")
preds_quota_df <- as.data.frame(preds_quota)
ggplot(preds_quota_df, aes(x = quota, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() +
  theme_minimal() +
  labs(title = "Predicted Prop Fem by Quota", y = "Proportion", x = "Quota")
```

### Proportion by Polyarchy

Predicted proportions across polyarchy range, holding quota at FALSE (no quota).

```{r}
#| label: preds-polyarchy
newdata_poly <- datagrid(model = model.beta,
                         polyarchy = seq(min(d.vdem$polyarchy, na.rm = TRUE), max(d.vdem$polyarchy, na.rm = TRUE), length.out = 50),
                         quota = FALSE)
preds_poly <- predictions(model.beta, newdata = newdata_poly)
ggplot(preds_poly, aes(x = polyarchy, y = estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  theme_minimal() +
  labs(title = "Predicted Prop Fem by Polyarchy (No Quota)", y = "Proportion", x = "Polyarchy")
```

### Explanation of Outcomes

The beta regression model indicates a significant positive effect of having a quota (estimate = 0.232, SE = 0.096, z = 2.417, p = 0.0156), suggesting that countries with gender quotas have a higher proportion of female MPs. Similarly, polyarchy has a significant positive effect (estimate = 0.008, SE = 0.002, z = 4.334, p < 0.001), implying that higher levels of democracy are associated with greater female representation in parliament.

The pseudo R-squared is 0.1073, indicating that the model explains about 10.7% of the variability in the data. The precision parameter phi is estimated at 12.01, which is significant.

Referring to the visualizations: The boxplot and density plots in Question 1 show that countries with quotas tend to have higher and more concentrated proportions of female MPs around higher values compared to those without. The point-range plot in Question 3 illustrates the model-predicted increase in proportion due to quotas, with non-overlapping confidence intervals suggesting a meaningful difference. The line plot in Question 4 demonstrates a gradual positive relationship between polyarchy and female representation, with the confidence band widening at extreme values due to fewer observations. Overall, these results highlight the roles of institutional (quotas) and systemic (democracy) factors in promoting gender equality in political representation.

Thank you.
