---
title: "Homework 2 - DSP1 - 2025/2026"
author: "Dushime Mudahera Richard 76250043"
toc: true
format: html
---

# Homework 2

Please use this file to complete your homework. This homework is meant to provide some practice working with data from importing through visualisation and to reflect on some core concepts in experimental design that we confront as analysts.

# Instructions

Step 1: Rename this file as follows: "DSP1-HW2-Name-StudentID" with your name and student ID replacing the appropriate parts.
Step 2: Update the author field in the YAML as indicated.
Step 3: Ensure the document has a Table of Contents when rendered (hint: you set this in the YAML)
Step 3: Complete to the Tasks below. Be sure to add meaningul label tags to each R code chunk and structure your analysis sensibly.
Step 4: Render to html or pdf (you may need to change target formats in the YAML)
Step 5: Zip (or otherwise compress) the html other relevant support folders and include this qmd file with your work filled in
Step 6: Submit the compressed folder which should be named "DSP1-HW2-Name-StudentID" through the e-classroom by Tuesday 25 Nov 23:00.

Hint: Make sure that when I open the html file I can view all the embedded images, see all your code and easily identify your responses to questions.

Note: Please see the rules for using AI on the e-classroom.

# Task 1 - Comparing two groups 

Determine if the weight of chickens eating two different diets differs using the chickwts dataset. Explore, analyse, and explain a statistical analysis comparing two groups. Include an explanation of each step you took and why you took that step.

Compare chicken weights between those eating meatmeal and soybean diets from the built-in chickwts dataset. Check the help for the dataset for a source reference.


```{r}
#| label: Summary of the dataset
#| echo: true
#| code-fold: true
#| code-summary: "View code"

summary(chickwts)
# view(chickwts)
```

## Loading and Filter the Data

Isolating the groups for my analysis, in this case i am interested in meatmeal and soybean feeds.

```{r}
#| label: load-filter
#| echo: true
#| code-fold: true
#| code-summary: "View code"

library(dplyr)

data(chickwts)
chick_two <- chickwts %>%
  filter(feed %in% c("meatmeal", "soybean"))
```

Summarizing chick_two 

```{r}
#| label: summarize
#| echo: true
#| code-fold: true
#| code-summary: "View code"

summary(chick_two)

chick_two %>%
  group_by(feed) %>%
  summarise(
    n = n(),
    mean_weight = mean(weight),
    sd_weight = sd(weight)
  )
```

## Visualizing Data with a Density Plot

Density show roughly normal distributions, similar variances, meatmeal is higher.

```{r}
#| label: vis-density-meatmeal-soybean
#| echo: true
#| code-fold: true
#| code-summary: "View code"

library(ggplot2)
ggplot(chick_two, aes(x = weight, fill = feed)) +
  geom_density(alpha = 0.6) +
  theme_classic() +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "Density Plot of Chicken Weights by Diet")
```
## Visualizing Data using a Boxplot

same here the boxplots show meatmeal is higher.

```{r}
#| label: vis-boxplot-meatmeal-soybean
#| echo: true
#| code-fold: true
#| code-summary: "View code"

ggplot(chick_two, aes(y = weight, x = feed, fill = feed)) +
  geom_boxplot() +
  theme_classic() +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "Boxplot of Chicken Weights by Diet")
```
Checking Normality (Shapiro-Wilk Test)

the assumption of normality for each group.

```{r}
#| label: check-normality-assumption
#| echo: true
#| code-fold: true
#| code-summary: "View code"

shapiro.test(chick_two$weight[chick_two$feed == "meatmeal"])
shapiro.test(chick_two$weight[chick_two$feed == "soybean"])

```
The Shapiro-Wilk test results show high p-values (0.9612 for meatmeal and 0.5064 for soybean), meaning we do not have enough evidence to reject the null hypothesis that the data is normally distributed, which satisfies the normality assumption required for a standard independent two-sample t-test.

## Check Equal Variances (F-Test)

```{r}
#| label: check-variance-(f-test)
#| echo: true
#| code-fold: true
#| code-summary: "View code"

var.test(weight ~ feed, data = chick_two)
```
The p-value of 0.5312 is much greater than the typical significance level of alpha = 0.05, leading us to fail to reject the null hypothesis of equal variances, thus satisfying the homogeneity of variance assumption required for the t-test (or allowing us to proceed with a pooled-variance t-test).


## T-test

t-test (assumptions met): t = 1.2525, df = 19.449, p-value = 0.2252

```{r}
#| label: T-test
#| echo: true
#| code-fold: true
#| code-summary: "View code"

t.test(weight ~ feed, data = chick_two)
```
We conclude that the observed difference in sample means meatmeal mean = 276.9091 g vs. soybean mean= 246.4286 g is not large enough to be considered statistically significant, and could reasonably be due to random chance or sampling variability.


## Visualisation of t-test plot

```{r}
#| label: vis-t-test-boxplot
#| echo: true
#| code-fold: true
#| code-summary: "View code"

ggplot(chick_two, aes(x = feed, y = weight, fill = feed)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, fill = "red") +
  theme_classic() +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "Boxplot with Means and Data Points for t-Test",
       y = "Weight (g)", x = "Feed Type")
```

# Task 2 - Comparing two methods of comparing two groups

Using the same dataset compare the diets of casein and linseed. Use both a t-test and a wilcoxon rank-sum test. Explain which one you would choose and show why you made that decision, then report the outcomes of that test.


Loading a new dataframe containing only the observations for the casein and linseed feed groups.

```{r}
#| label: load-filter-casein-linseed
#| echo: true
#| code-fold: true
#| code-summary: "View code"

data(chickwts)
chick_case_lin <- chickwts %>%
  filter(feed %in% c("casein", "linseed"))
```


```{r}
#| label: summary-stats
#| echo: true
#| code-fold: true
#| code-summary: "View code"

chick_case_lin %>%
  group_by(feed) %>%
  summarise(
    n = n(),
    mean_weight = mean(weight),
    sd_weight = sd(weight)
  )
```

Visualizing Data with a Density Plot

```{r}
#| label: vis-density
#| echo: true
#| code-fold: true
#| code-summary: "View code"

ggplot(chick_case_lin, aes(x = weight, fill = feed)) +
  geom_density(alpha = 0.6) +
  theme_classic() +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "Density Plot of Chicken Weights: Casein vs Linseed")
```
The casein appears a little higher 

Normality Check 

```{r}
#| label: check-normality
#| echo: true
#| code-fold: true
#| code-summary: "View code"

shapiro.test(chick_case_lin$weight[chick_case_lin$feed == "casein"])
shapiro.test(chick_case_lin$weight[chick_case_lin$feed == "linseed"])
```

The Shapiro-Wilk test results show p-values (0.2592 for casein and 0.9035 for linseed), meaning we do not have enough evidence to reject the null hypothesis that the data is normally distributed for either group, which satisfies the normality assumption required for a standard independent two-sample t-test.


```{r}
#| label: check-variance
#| echo: true
#| code-fold: true
#| code-summary: "View code"

var.test(weight ~ feed, data = chick_case_lin)
```
The p-value of 0.4977 is much greater than the typical significance level of alpha = 0.05, leading us to fail to reject the null hypothesis of equal variances, thus confirming that the homogeneity of variance assumption is met and a standard t-test would be appropriate regarding this assumption.



```{r}
#| label: t-test
#| echo: true
#| code-fold: true
#| code-summary: "View code"

t.test(weight ~ feed, data = chick_case_lin)
```
Since the p-value is 0.0002606, which is much smaller than the conventional significance level of alpha = 0.05, we reject the null hypothesis.

concluding  that the mean weight of chickens fed casein is significantly different from the mean weight of chickens fed linseed.

There is a high difference between the 2 



```{r}
#| label: wilcoxon
#| echo: true
#| code-fold: true
#| code-summary: "View code"

wilcox.test(weight ~ feed, data = chick_case_lin, conf.int = TRUE)
```
While I dont have too much argument on which to choose and why, I would go choose randomly based on the  values i am getting 
Both the normality assumption (based on Shapiro-Wilk p-values) and the homogeneity of variance assumption (based on the F-test p-value) were met. I would choose the t-test.

Reference:

https://medium.com/@sujathamudadla1213/what-is-the-main-difference-between-the-t-test-and-the-wilcoxon-test-in-terms-of-their-assumptions-ebff4c8385aa 


# Task 3 - Linear regression - Iris data

Fit a linear regression predicting petal width based on petal length and species of iris using the built-in iris dataset

Answer the following questions/complete the following tasks:

a. Plot the data in a useful way to show any potential relationships. Add a reference to the data source as a caption to the plot (hint: check help(iris) and use the Anderson reference).

```{r}
#| label: Iris plot
#| echo: true
#| code-fold: true
#| code-summary: "View code"

data(iris)
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point() +
  theme_classic() +
  labs(title = "Petal Width vs Petal Length by Iris Species",
       caption = "Anderson, E. (1935). The irises of the Gaspé Peninsula. Bulletin of the American Iris Society, 59, 2–5.")
```


b. Fit a linear regression to the data. Is linear regression appropriate for the data?

```{r}
#| label: fit-lm-iris
#| echo: true
#| code-fold: true
#| code-summary: "View code"

iris_lm <- lm(Petal.Width ~ Petal.Length * Species, data = iris)
summary(iris_lm)

# Diagnostic plots to check appropriateness
plot(iris_lm)
```
yes, the model provides a strong predictive fit (R^2=0.9477),


c. What is the effect of a one unit increase in petal length on petal width?

```{r}
#| label: coefficients-iris
#| echo: true
#| code-fold: true
#| code-summary: "View code"

coef(iris_lm)
```
The effect of a one-unit increase in Petal Length on Petal Width is not constant but varies by species,


d. Are there statistically significant differences in petal width between species? 
  - Create a table with the difference that includes an estimate of the difference and confidence intervals for those estimates.
  - Create a visualisation of the differences
  - Write a few sentences that describe the differences using the estimates and CIs you have found.

 Differences Between Species (Using emmeans)
```{r}
#| label: emmeans-iris
#| echo: true
#| code-fold: true
#| code-summary: "View code"

library(emmeans)

iris_em <- emmeans(iris_lm, ~ Species | Petal.Length, at = list(Petal.Length = mean(iris$Petal.Length)))
iris_pairs <- pairs(iris_em)
confint(iris_pairs)

# Table (as tibble)
as_tibble(confint(iris_pairs))
```
Yes, there are statistically significant differences in Petal Width for two out of the three species pairs and its Setosa vs. Virginica 

The difference between Setosa and Virginica is significant, with Virginica having an estimated Petal Width 1.03 cm larger (95% CI: -1.8513922 to -0.2093448 ) than Setosa. 

Similarly, Versicolor has a significantly smaller Petal Width than Virginica, with an estimated difference of 0.5786162 cm (95% CI: -0.8003913 to -0.3568410). However, the estimated difference between Setosa and Versicolor (-0.4517524 cm) is not statistically significant as its 95% confidence interval [-1.25, 0.348] includes zero.

Below is a plot


Visualization of Differences
```{r}
#| label: vis-differences-iris
#| echo: true
#| code-fold: true
#| code-summary: "View code"

library(tidyverse)
iris_pairs_df <- as_tibble(confint(iris_pairs))
ggplot(iris_pairs_df, aes(x = contrast, y = estimate, ymin = lower.CL, ymax = upper.CL)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Pairwise Differences in Petal Width Between Species",
       y = "Difference in Petal Width (mm)", x = "Contrast")
```



e. Create a visualisation of the petal width data with the regression lines for each species overlayed on the points.

```{r}
#| label: vis-reg-lines-iris
#| echo: true
#| code-fold: true
#| code-summary: "View code"

ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() +
  labs(title = "Petal Width vs Petal Length with Regression Lines by Species")
```

