---
title: "DSP 1 - Final 2025/2026"
author: "RDM"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
format:
  html:
    toc: true
    toc-depth: 2
---

# Instructions

1. Update the YAML Header with the information requested. Add a table of contents with depth 2.
2. Change the filename to DPS1_Final_2025-2026_Lastname-studentID.(qmd and html)
3. Complete the tasks as described below.
  * If you use any packages we haven't used in class, please explain why you chose them
4. Use good coding practices, label your code chunks, provide explanatory comments, ensure it is easily readable. 
5. Observe the rules for AI. If you use it provide specific descriptions of how it was used as requested on the e-classroom, not only a general comment.
6. Submit the .qmd file and .html file together through the e-classroom.
  * The html file may be self-contained (check online for the yaml instruction) or submitted with the dependent folders/files. Make sure the file you submit can be opened and viewed in complete form with the files you submit.

# Setup

```{r}
#| label: setup
#| message: false
#| warning: false

# Load necessary libraries
library(tidyverse) # Core data manipulation and visualization (ggplot2, dplyr, readr)
library(fs)        # File system operations (handling file paths in Task 3)
library(broom)     # Tidy model output (converting model objects to dataframes)
library(gt)        # Grammar of Tables (creating professional tables)
library(emmeans)   # Estimated Marginal Means (post-hoc comparisons for regression)

# Set a consistent theme for all plots
# theme_minimal is clean and uses a white background, suitable for reports
theme_set(theme_minimal())
```

# Task 1 - Visualising distributions

Create a visualisation that displays the four distributions described below with the requested annotations. Use a theme other than the default.

Distributions: 
  1. Beta(3,12)
  2. Gamma(5,10)
  3. Lognormal(5,10)
  4. Normal(5,5) 
  
For each distribution:

  1. Fill in the area between the 33rd and 66th quantiles (i.e., 0.33, 0.66) so that the range is clearly differentiated from the rest of the distribution.
  2. Mark the mean and median on the chart (make sure it is clear which mark is the mean and which is the median)
  3. Annotate with the mean and median, rounded to 2 decimal points.
  
The charts should be combined into a single visualisation. The x-axis limits should make sure the distribution is clearly defined and easy to interpret.

**Bonus - 5 extra points**: Recreate the charts from Task 1 and add the mode of each distribution.

*Included the bonus in the implementation below.*

```{r}
#| label: task1-vis
#| fig.width: 10
#| fig.height: 8

# Define properties for each distribution
# Used lists to store parameters for flexibility

# Function to convert linear mean/sd to lognormal parameters
# R's dlnorm uses meanlog and sdlog (parameters of the underlying normal distribution).
# If we assume the instructions "Lognormal(5,10)" refer to the Linear Mean=5 and Linear SD=10,
# we must convert them using the Method of Moments.
get_lognorm_params <- function(mean, sd) {
  list(
    # Formula for meanlog: log(m^2 / sqrt(v + m^2))
    meanlog = log(mean^2 / sqrt(sd^2 + mean^2)),
    # Formula for sdlog: sqrt(log(1 + v/m^2))
    sdlog = sqrt(log(1 + (sd^2 / mean^2)))
  )
}

# Apply conversion for Lognormal(5,10)
ln_params <- get_lognorm_params(5, 10)

# List of distributions with their specific functions and parameters
# This allows us to iterate over them generically
dists <- list(
  Beta = list(
    fun = dbeta,   # Density function
    params = list(shape1 = 3, shape2 = 12),
    qfun = qbeta,  # Quantile function (for shading)
    range = c(0, 1), # Support of Beta is [0,1]
    stats = c(
      mean = 3/(3+12), # Expected value: a / (a+b)
      mode = (3-1)/(3+12-2) # Mode: (a-1)/(a+b-2)
    )
  ),
  Gamma = list(
    fun = dgamma,
    # Assuming parameters are Shape (k)=5 and Rate=10 based on standard notation often used in context.
    # (If Scale was meant, params would constitute a much wider distribution).
    params = list(shape = 5, rate = 10), 
    qfun = qgamma,
    # Custom range for better visibility (zoomed in)
    manual_range = c(0, 1.5),
    stats = c(
      mean = 5/10, # Mean: shape / rate
      mode = (5-1)/10 # Mode: (shape-1) / rate
    )
  ),
  Lognormal = list(
    fun = dlnorm,
    params = ln_params, # Calculated above
    qfun = qlnorm,
    # Custom range to show the mode (near 0) and the long tail up to 25
    manual_range = c(0, 25), 
    stats = c(
      mean = 5, # As defined
      # Mode of lognormal: exp(meanlog - sdlog^2)
      mode = exp(ln_params$meanlog - ln_params$sdlog^2)
    )
  ),
  Normal = list(
    fun = dnorm,
    params = list(mean = 5, sd = 5),
    qfun = qnorm,
    # Custom range: Starts at -10 to show the full left tail clearly
    manual_range = c(-10, 20),
    stats = c(
      mean = 5,
      mode = 5 # For Normal, Mean = Median = Mode
    )
  )
)

# Helper function to generate data frame for plotting
# Arguments: 'name' (dist name), 'info' (list from dists above)
generate_dist_data <- function(name, info) {
  # determine x range
  # Use manual range if provided, otherwise calculate from quantiles (0.1% to 99.9%)
  if (!is.null(info$manual_range)) {
    x <- seq(info$manual_range[1], info$manual_range[2], length.out = 1000)
  } else {
    lb <- info$qfun(0.001, info$params[[1]], info$params[[2]])
    ub <- info$qfun(0.999, info$params[[1]], info$params[[2]])
    # Ensure positive distributions don't go below 0
    if(name %in% c("Beta", "Gamma", "Lognormal")) lb <- max(0, lb)
    x <- seq(lb, ub, length.out = 1000)
  }
  
  # Calculate density (y-values) using do.call to pass list of params to function
  y <- do.call(info$fun, c(list(x = x), info$params))
  
  # Calculate 33rd and 66th percentiles for shading
  q33 <- do.call(info$qfun, c(list(p = 0.33), info$params))
  q66 <- do.call(info$qfun, c(list(p = 0.66), info$params))
  
  # Calculate exact Median
  median_val <- do.call(info$qfun, c(list(p = 0.5), info$params))
  
  # Extract pre-calculated Mean and Mode
  mean_val <- info$stats["mean"]
  mode_val <- info$stats["mode"]
  
  # Return a tidy dataframe
  tibble(
    Distribution = name,
    x = x,
    y = y,
    q33 = q33,
    q66 = q66,
    mean = mean_val,
    median = median_val,
    mode = mode_val
  )
}

# Combine all data into one dataframe using map_df
plot_data <- map_df(names(dists), ~generate_dist_data(., dists[[.]]))

# Plot Task 1 (Mean + Median only)
# Using ggplot2 for visualization
p_task1 <- ggplot(plot_data, aes(x, y)) +
  # Line for the distribution curve
  geom_line(color = "black") +
  # Area shading between 33rd and 66th percentiles
  # We filter the data to only include x values within this range
  geom_area(data = plot_data %>% filter(x >= q33 & x <= q66), fill = "skyblue", alpha = 0.5) +
  # Vertical line for Mean (Red Dashed)
  geom_vline(aes(xintercept = mean), color = "red", linetype = "dashed") +
  # Vertical line for Median (Blue Dotted)
  geom_vline(aes(xintercept = median), color = "blue", linetype = "dotted") +
  # Text annotations: Placed at the top (y=Inf)
  geom_text(data = plot_data %>% distinct(Distribution, mean, median),
            aes(x = median, y = Inf, 
                label = paste0("Mean: ", round(mean, 2), "\nMedian: ", round(median, 2))),
            vjust = 1.5, size = 3, color = "black") +
  # Facet wrap creates separate panels for each distribution
  # scales="free" is critical because ranges and heights vary significantly
  facet_wrap(~Distribution, scales = "free") +
  labs(
    title = "Probability Distributions",
    subtitle = "Shaded: 33rd-66th percentiles. Red: Mean, Blue: Median",
    x = "Value",
    y = "Density"
  ) +
  theme(strip.text = element_text(face = "bold", size = 12))

p_task1

```

---

# Task 2 - Fit a regression and report the outcomes

Fit a regression to the data from fibre_reinforced.csv with str as the dependent variable with density and treatment as the independent variables. The data is from a study investigating using flax fibre cord treated with different materials as reinforcement in concrete.

1. Visualise the raw data
2. Choose an appropriate regression method and fit the regression
3. Visualise the regression outputs, annotating the visualisation with relevant information.
4. Use a table layout package (i.e., tinytables, gt, kable) to display the regression outputs.
5. Write a summary of the findings, defining effect sizes. In this case an effect size of more than 10

```{r}
#| label: task2-load
#| message: false

# Load Data
# read_csv is faster and more robust than read.csv
fibre_data <- read_csv("data/fibre_reinforced.csv")
```

### 1. Visualise the raw data

```{r}
#| label: task2-vis-raw
#| layout-ncol: 2 
# layout-ncol: 2 forces the two plots (p1 and p2) to appear side-by-side in the output

# Boxplot of Strength by Treatment
# Shows distribution of strength for each categorical treatment type
p1 <- ggplot(fibre_data, aes(x = treatment, y = str, fill = treatment)) +
  geom_boxplot() +
  labs(title = "Strength by Treatment", y = "Strength (str)", x = "Treatment") +
  theme(legend.position = "none")

# Scatter of Strength vs Density
# Shows relationship between continuous variables, colored by treatment
p2 <- ggplot(fibre_data, aes(x = density, y = str, color = treatment)) +
  geom_point() +
  # Add linear trend lines to see if density typically increases strength
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Strength vs Density", y = "Strength (str)", x = "Density")

# Print plots (will be displayed side-by-side due to layout-ncol)
p1
p2
```

### 2. Fit Regression

Chosen method: **Linear Regression (`lm`)**. 
**Justification**: The dependent variable `str` (strength) is continuous. The predictors include a continuous variable (`density`) and a categorical variable (`treatment`). A linear model is appropriate to estimate the effect of treatment while controlling for density, assuming linearity and normality of residuals (which can be checked).

```{r}
#| label: task2-model

# Fit linear model
# Formula: str (Dependent) ~ density (Indep) + treatment (Indep)
lm_model <- lm(str ~ density + treatment, data = fibre_data)

# Summary of the model (R-squared, coefficients, F-statistic)
summary(lm_model)
```

### 3. Visualise Regression Outputs

```{r}
#| label: task2-vis-results

# Extract results with broom::tidy
# conf.int = TRUE calculates the 95% Confidence Intervals for us
tidy_results <- tidy(lm_model, conf.int = TRUE)

# Coefficient Plot (Forest Plot)
# Visualizes the estimate and its uncertainty
ggplot(tidy_results %>% filter(term != "(Intercept)"), 
       aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high)) +
  geom_pointrange(color = "darkblue") + # Point for estimate, line for CI
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # Ref line at 0 (null hypothesis)
  labs(
    title = "Regression Coefficients (95% CI)",
    subtitle = "Effect of Density and Treatment on Strength",
    x = "Estimate",
    y = "Term"
  ) +
  geom_text(aes(label = round(estimate, 2)), vjust = -1)
```

### 4. Tables

**Table 1: Regression Coefficients**

```{r}
#| label: task2-table-coefs

# Create a professional table using 'gt'
tidy_results %>%
  gt() %>%
  tab_header(title = "Linear Regression Results", subtitle = "Dependent Variable: Strength") %>%
  # Format numbers to 2 decimal places for readability
  fmt_number(columns = c(estimate, std.error, statistic, conf.low, conf.high), decimals = 2) %>%
  # Format p-values scientifically (e.g., 2e-16)
  fmt_scientific(columns = p.value)
```

**Table 2: Comparisons between Treatments**

```{r}
#| label: task2-table-emmeans

# Pairwise comparisons using Estimated Marginal Means (emmeans)
# This compares each treatment against every other treatment, averaged over density
em_means <- emmeans(lm_model, pairwise ~ treatment)
em_contrasts <- as.data.frame(em_means$contrasts)

# Display contrasts in a table
em_contrasts %>%
  gt() %>%
  tab_header(title = "Pairwise Comparisons of Treatments") %>%
  fmt_number(columns = c(estimate, SE, t.ratio), decimals = 2) %>%
  fmt_scientific(columns = p.value)
```

### 5. Summary of Findings

```{r}
#| label: task2-summary
#| echo: false
#| results: asis 
#| # results: asis ensures that cat() output is rendered as Markdown text, not code output

# Extracting significant effects > 10 (absolute value)
sig_effects <- tidy_results %>% 
  filter(abs(estimate) > 10 & term != "(Intercept)")

cat("**Key Findings:**\n\n")
cat("The regression analysis allows us to control for density while comparing treatments.\n\n")

# Dynamic text generation based on model results
if(nrow(sig_effects) > 0) {
  cat("The following variables had an effect size greater than 10:\n")
  for(i in 1:nrow(sig_effects)) {
    # Print bullet point for each large effect
    cat(paste0("- **", sig_effects$term[i], "**: Estimate = ", round(sig_effects$estimate[i], 2), "\n"))
  }
} else {
  cat("No variables showed an effect size greater than 10 in the model coefficients.\n")
}

cat("\n(See Table 2 for pairwise differences between treatments).")

```

---

# Task 3 - Read and combine files using purrr:map and fs

Using a combination of the fs package and purrr package (specifically using a map function - choose the appropriate function) to read the contents of multiple files into a dataframe or tibble. The files are in the thermal_comfort folder.

When reading the files, add columns that contain the two components extracted from the filename (the portion that does not include the extension or any path components): condition and date, separately. The file name identifies the condition id, the date of the measurements in the form "ConditionID-Date.csv". Read the "conditions.csv" file separately, and merge the specific conditions into the combined dataframe or tibble you produced using a join function.

The final merged and joined file should have the following columns: date, subject_id, condition_id, rating, material_ratio, temperature in that order.

After reading in the file, make a visualisation and summary table (using, e.g., tinytable, gt, or kable) that represents the data appropriately grouping it by room condition (material ratio and temperature).

```{r}
#| label: task3-process

# 1. Get file paths
# fs::dir_ls is convenient for getting all files matching a pattern
file_paths <- dir_ls("data/thermal_comfort", glob = "*.csv")

# 2. Function to read and annotate
# This function will be applied to each file
read_and_annotate <- function(path) {
  # Extract filename without extension (e.g., "A01-20251010")
  fname <- path_file(path) %>% path_ext_remove()
  
  # Split 'ConditionID-Date' string into parts
  parts <- str_split(fname, "-")[[1]]
  cond_id <- parts[1]
  date_val <- parts[2]
  
  # Read CSV
  # Use show_col_types = FALSE to silence messages during loop
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      condition_id = cond_id,
      date = date_val
    )
}

# 3. Process all files with map_dfr
# map_dfr iterates over 'file_paths', runs 'read_and_annotate', and binds results into one dataframe
combined_ratings <- map_dfr(file_paths, read_and_annotate)

# 4. Read Loop Conditions
# Explicitly specifying col_types to avoid parsing errors
# (e.g., preventing "30:70" from being misread as time)
conditions_df <- read_csv("data/conditions.csv", 
                          col_types = cols(
                            material_ratio = col_character(),
                            temperature = col_double(),
                            code = col_character()
                          ))

# 5. Join Code
# Merge the ratings with the condition details
final_df <- combined_ratings %>%
  # Join on the condition ID (called 'condition_id' in ratings, 'code' in conditions)
  left_join(conditions_df, by = c("condition_id" = "code")) %>%
  select(date, subject_id, condition_id, rating, material_ratio, temperature)

# Show structure to verify results
glimpse(final_df)
```

### Visualisation and Summary

```{r}
#| label: task3-vis-summary

# Summary Table
summary_stats <- final_df %>%
  group_by(material_ratio, temperature) %>%
  summarise(
    Mean_Rating = mean(rating, na.rm = TRUE),
    Median_Rating = median(rating, na.rm = TRUE),
    Count = n(),
    .groups = "drop" # Drop grouping after summary
  )

summary_stats %>%
  gt() %>%
  tab_header(title = "Thermal Comfort Ratings by Condition") %>%
  fmt_number(columns = c(Mean_Rating), decimals = 2)

# Visualization
ggplot(final_df, aes(x = material_ratio, y = rating, fill = factor(temperature))) +
  geom_boxplot() +
  labs(
    title = "Thermal Comfort Ratings",
    x = "Material Ratio",
    y = "Rating (1-7)",
    fill = "Temperature"
  ) +
  theme_minimal()
```


# Bonus - 5 extra points

Recreate the charts from Task 1 and add the mode of each distribution.

```{r}
#| label: bonus-vis
#| fig.width: 10
#| fig.height: 8

# Re-using plot_data from Task 1
# This code block is almost identical to Task 1, but includes the Mode line.
ggplot(plot_data, aes(x, y)) +
  geom_line(color = "black") +
  # Area shading
  geom_area(data = plot_data %>% filter(x >= q33 & x <= q66), fill = "skyblue", alpha = 0.5) +
  # Mean line (Red, Dashed)
  geom_vline(aes(xintercept = mean), color = "red", linetype = "dashed") +
  # Median line (Blue, Dotted)
  geom_vline(aes(xintercept = median), color = "blue", linetype = "dotted") +
  # Mode line (Green, Solid) - Bonus inclusion
  geom_vline(aes(xintercept = mode), color = "darkgreen", linetype = "solid") +
  # Text annotations including Mode
  geom_text(data = plot_data %>% distinct(Distribution, mean, median, mode),
            aes(x = median, y = Inf, 
                label = paste0("Mean: ", round(mean, 2), "\nMedian: ", round(median, 2), "\nMode: ", round(mode, 2))),
            vjust = 1.5, size = 3, color = "black") +
  facet_wrap(~Distribution, scales = "free") +
  labs(
    title = "Probability Distributions (with Mode)",
    subtitle = "Shaded: 33rd-66th percentiles. Red: Mean, Blue: Median, Green: Mode",
    x = "Value",
    y = "Density"
  ) +
  theme(strip.text = element_text(face = "bold", size = 12))
```

# Thank you
