---
title: "DSP I - 2025/2026 - Beta regression - Bayesian style"
author: "Mike B."
format: revealjs
---

## Goal: model survey response data on a likert-like scale

Data on a likert-like scale are categories in a specific order. They seek to measure something that cant be measured directly (like beauty, or interestingness, or agreement). This is called a latent variable. (There are other ways to model latent variables, too).

## Motivation

A survey (ongoing!) asks respondents to rate pictures of walls along several likert-like scales, including boring-interesting (this is a semantic differential scale, which has opposite adjectives on either end of the scale).

We will examine if boring-interesting ratings differ between wall types.

## Our setup

```{r}
#| label: setup

library(tidyverse)
library(betareg)
library(brms)
library(distributional)
library(ggdist)
library(tidybayes)
library(bayestestR)

palette <- viridis::viridis(10, option = "turbo")

`%notin%` <- Negate(`%in%`)
inv.logit <- function(x) { exp(x) / (1 + exp(x))}
```

## Data

```{r}
#| label: data


d.sound <- read_csv("data/sound_acceptability.csv")

# summary(d.sound)
# head(d.sound)


```

## Visualise

```{r}
#| label: visualise

ggplot(data = d.sound, aes(x=accept.mu)) + 
  geom_density()


ggplot(data = d.sound, aes(x=accept.mu, y= screed.alpha)) + 
  stat_halfeye(point_interval = "median_hdi")

ggplot(data = d.sound, aes(x=accept.mu, y= screed.alpha)) + 
  stat_halfeye(point_interval = "median_qi")
```

## Analyse - Beta regression

Beta regression is used when we have data bounded between in the range 0 < y < 1. The beta distribution is not defined for y = 0 or y = 1, so we refer to these as degenerate values. If we have zeroes and ones in our data set we have a couple options:

  1. Fit a Zero-One inflated beta regression model or ordered beta regression
  2. With very few 0s or 1s we may consider dropping those values or adding or subtracting a small quantity (i.e., 0.0001) to/from the degenerate values.

We aren't limited to values that are explicitly between 0 and 1. Scaling to values between 0 and 1 is quite easy, so if we have a scale with with five scored items between 1 and 5, we have a minimum score of 5 and a maximum score of 25. We can rescale this range to 0,1 fit our beta model and then scale the results back to the original scale quite easily. (This is a case where degenerate values are likely to occur, so option 1 above is useful.)

The beta distribution has two characteristics: shape (alpha) and scale (beta).

```{r}
#| label: b

dbeta <- tibble(
  shape = c(1, 1, 5, 5, 10, 10),
  scale= rep(c(5,10), times=3)
)

ggplot(data = dbeta, aes(xdist= dist_beta(scale, shape),
                         colour = as.factor(shape),
                         linetype = as.factor(scale)
                         ) ) +

  stat_slab(fill=NA) +
  coord_cartesian(expand = FALSE) 
  #scale_colour_viridis_b()

```

 
Why not linear regression? Linear regressions assumes unbounded ranges. Even with low probabilities of values outside the observable range, out estimates are likely to be biased to some degree.

Our data doesn't currently have a column with values 0 < y < 1, so we should create one:

```{r}

glimpse(d.sound)
d.sound$accept.beta <- d.sound$accept.mu / 10


```

## Bayesian regression



## Priors

To fit a Bayesian regression model we need to know another piece of information that we haven't encountered before: _Priors_. Priors are our prior knowledge - what we know going into our analysis. While this can be intuition, it best to utilise information from previously gained knowledge or to use _uninformative_ or _weekly informative_ priors. 

We use priors to describe an expected range for the estimated betas, intercepts, and variances to take. These require a some understanding of the scale the betas are likely to be on. For example, fitting a standard (non-transformed) linear regression, the beta coefficient for a slope will be the mean value from the average for the baseline case and could be -120, or zero, or 146 if that's the sample mean. However, in regressions that use other scales like a logit scale or log scale) large values are exceedingly rare. We need to set our priors accordingly.

```{r}
#| label: priors

default_prior(accept.beta ~ screed.alpha + speed + shoes , data= d.sound , family ="beta" )

priors <- c(
  prior("normal(0,3)", class = "b")
)
```

By default, we have a "flat" prior for our betas, and student's t distribution for the intercept and sigma. A flat distribution is a uniform 1 distribution.


```{r}
#| label: model_fitting

sound.beta <- brm(
  accept.beta ~ screed.alpha + shoes + speed,
  data = d.sound,
  family = "beta",
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 1000,
  seed = 1234
)

```


```{r}
#| label: model_checking
summary(sound.beta)


```

```{r}
#| label: extracting_estimates


```

```{r}
#| label: visualising_estimates


```
