---
title: "Logistic regression"
author: "mike burnard"
date: "17 Nov 2025"
format: revealjs
---

```{r}
#| label: setup
#| include: false

#standard set
library(tidyverse)
library(emmeans)
library(marginaleffects)

#introducing
# you may need to install these with install.packages()
library(broom)
library(tidymodels)
#library(ISLR)
```

## Logistic regression

- Used when we have a binary outcome (0,1 | Yes,No), etc.
- A _classification_ method
- In R, the process is similar to linear regression

Example: Your bank wants to assess if online transactions are fraud or not. They might look at IP address, website, transaction amounts as independent variables, but the dependent variable is binary: Fraud or Authentic.

We will look at another case: given annual income, and loan balance can we predict if someone will default on their loan (fail to make the payment).

## Training and testing datasets

When we fit a model to experimentally collected data - when there was a planned experiment - we are using statistical models to explain the data and (sometimes) make predictions future experiments based on the data.

In a machine/statistical learning environment we are often trying to fit a model to a large set of data to predict future events. The same techniques are commonly applied. However, in ML we often use a training dataset to fit the model and testing dataset to assess how well the model fits. 

In this case, we will take use the Default dataset from the ISLR package, and split it into a training data set with 60 % of the values (~6k), and a testing data set with the rest (~4k)

## Data

```{r}
#| label: data
#| echo: true
#| code-fold: true
#| code-summary: "click for code"


```


## Visualise

```{r }
#| label: vis_raw1
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Visualisations continuned

```{r }
#| label: vis_raw2
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Fitting logistic regression

```{r }
#| label: model 1
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```
## Extracting values from coefficients

```{r}
#| label: coefs
#| echo: true
#| code-fold: true
#| code-summary: "click for code"

```



## Comparing models with ANOVA

```{r}
#| label: model_compare
#| echo: true
#| code-fold: true
#| code-summary: "click for code"



```
## Extracting information with broom

```{r}
#| label: extract_with_broom
#| echo: true
#| code-fold: true
#| code-summary: "click for code"


```


## Comparisons

```{r}
#| label: comparisons
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```


## Predictions

```{r }
#| label: preds
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Vis predictions

```{r}
#| label: vis_preds
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```

## Assessing the model

```{r }
#| label: mcfaddans_psr2
#| echo: true
#| code-fold: true
#| code-summary: "click for code"

```

## Classification rate

Here we use the test data we created earlier to see how well our model performs on the other part of the original dataset.

```{r }
#| label: class_rat
#| echo: true
#| code-fold: true
#| code-summary: "click for code"
```