---
title: "penguins dataset"
format: html
---

```{r}
# view(penguins)
glimpse(penguins)
```

```{r}
ggplot(data = penguins,
       mapping = aes(x=flipper_len, y=body_mass)
) +
  geom_point(mapping = aes(color=species, shape = species)) + 
  geom_smooth(method = "lm") + 
  labs(
    title = "Body mass and Flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length in (mm)",
    y = "Body mass in (g)",
    color = "Species",
    shape = "Species"
  )
  # scale_color_colorblind()

```

```{r}
ggplot(
  data = penguins,
  aes(x=bill_len, y=bill_dep)
) + 
  geom_point(aes(color=species, shape=species)) + 
  geom_smooth(method = "lm")
```

```{r}
ggplot(data = penguins,
       aes(x=species, y=bill_dep, color=island)) + 
  geom_point()
```


```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_len, y = body_mass, color=species)
) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "First"
  )

ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_len, y = body_mass,colour =species )
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_len, y = body_mass)
  ) + 
  labs(
    title = "Second"
  )
```

# ggplot2 calls

```{r}
ggplot(penguins, aes(x=flipper_len, y= body_mass)
    ) + 
  geom_point() 
  

```

using the pipe, |>:

```{r}
penguins |> 
  ggplot(aes(x = flipper_len, y = body_mass, colour = species)) + 
  geom_point()

```
# bar charts 
for categorical variables 

```{r}
ggplot(penguins, aes(x=fct_infreq(species), fill = species)
       ) + 
  geom_bar()

# relationship btn island and species 

ggplot(penguins, aes(x=island , fill = species)
       ) + 
  geom_bar(position = "fill") + 
  labs(
    y = "proportion"
  )
```

# hiistograms 
good for numercial variables 

```{r}
ggplot(penguins, aes(x=body_mass)
       ) + 
  geom_histogram(binwidth = 200)
```


# desnsity plot 

A density plot is a smoothed-out version of a histogram and a practical alternative, particularly for continuous data that comes from an underlying smooth distribution

```{r}
ggplot(penguins, aes(x=body_mass)
       ) + 
  geom_density()
```

# Visualizing relationships

box plot 

```{r}
ggplot(penguins, aes(x=species, y=body_mass)
       ) + 
  geom_boxplot()
```

```{r}
ggplot(penguins, aes(x=body_mass, colour = species, fill = species)
       ) + 
  geom_density(linewidth = 0.75, alpha=0.2)

```

# 3 or more variables

```{r}
ggplot(penguins, aes(x = flipper_len, y = body_mass)) +
  geom_point(aes(color = species, shape = island))
```
facet wrap is useful for categorical variables 

```{r}
ggplot(penguins, aes(x = flipper_len, y = body_mass)) +
  geom_point(aes(color = species, shape = species)) +
  facet_wrap(~island)

ggsave(filename = "penguins_analysis.png")
```

# MPG dataset 
The mpg data frame contains 234 observations collected by the US Environmental Protection Agency on 38 car models

```{r}
glimpse(mpg)
```

## visualization

```{r}
ggplot(mpg, aes(x = fct_infreq(class), fill = class)) +
  geom_bar()

ggplot(mpg, aes(x = cty, y = hwy, color=class)) +
  geom_point()

ggsave("mpg-plot.png")
```

```{r}
ggplot(mpg, aes(x= cty, color=class)) + 
  geom_density()
```


